Using Linear Regression to Examine the Correlation Between Access to Drinking Water and Life Expectancy

### 2022-12-09 Manuscript/Paper.tex
additions to discussion

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage[margin=1 in]{geometry}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{natbib}
\usepackage{lipsum}
\usepackage[colorlinks=true, citecolor=blue]{hyperref}
\usepackage{listings}
\usepackage{changepage}
\setlength{\parindent}{0pt}

\title{Using Linear Regression to Examine the Correlation Between Access to Drinking Water and Life Expectancy}
\author{Julia Andronowitz\\
University of Connecticut\\Statistics Department}
\date{November 14, 2022}

@@

\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
import numpy as np
import pandas as pd
import seaborn as sns
from bokeh.plotting import figure
from bokeh.io import show, output_notebook
from bokeh.models import ColumnDataSource, HoverTool
from bokeh.layouts import gridplot
output_notebook()
\end{lstlisting}
\end{adjustwidth}

Next, we import the data:
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
rawData1=pd.read_csv('archive/basicDrinkingWaterServices.csv')
rawData2=pd.read_csv('archive/lifeExpectancyAtBirth.csv')
\end{lstlisting}
\end{adjustwidth}

If we would like to see the data, use the command
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
rawData1
\end{lstlisting}
\end{adjustwidth}

An output of the data is below:

\begin{center}
\includegraphics[width=5in]{figure1.pdf}
\includegraphics[width=5in]{figure2.pdf}
\end{center}

For the drinking water data, we take only the columns with the country, year, and the percentage of the population using at least basic drinking water services.
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
rawData1=rawData1[["Location", "Period" , "First Tooltip"]]
\end{lstlisting}
\end{adjustwidth}

We do the same for the life expectancy data, only taking the columns with the country, year, sex, and average life expectancy at birth.
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
rawData2=rawData2[["Location", "Period", "Dim1", "First Tooltip"]]
\end{lstlisting}
\end{adjustwidth}

Next, we want to convert our pandas dataframes into numpy arrays so that we can use the appropriate mathematical functions.
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
data1=np.array(rawData1)
data2=np.array(rawData2)
\end{lstlisting}
\end{adjustwidth}

We want to have as many constants as possible in our model. Since the data was collected starting in 2000, we must use the same years in both data sets. We will use the year 2015 as the benchmark. The data can then be filtered using a loop and appending the 2015 data to a new list.
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
dw2015=[]
for x in data1:
    if x[1]==2015:
        dw2015.append(x)
\end{lstlisting}
\end{adjustwidth}

This must be done for both data sets. We now do it for the life expectancy data set. As previously mentioned in the "Data" section, we also want to use only the rows with both sexes.
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
le2015=[]
for x in data2:
    if (x[1]==2015) and (x[2]=='Both sexes'):
        le2015.append(x)
\end{lstlisting}
\end{adjustwidth}

These lists will also be converted into numpy arrays.
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
DW2015=np.array(dw2015)
lifeExpectancy2015=np.array(le2015)
\end{lstlisting}
\end{adjustwidth}

Now, we see the shapes of both these arrays are different. The drinking water array contains 193 rows and 3 columns while the life expectancy array contains 183 rows and 4 columns. This means that there are some countries that are not in both lists. So, we need to delete the data for countries that are only in one or the other. We do this by checking if the countries in the drinking water array are also in the life expectancy data. If both countries are in both data sets, we append them to our new drinking water list, which is what we will use going forward.
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
drinkingWater2015=[]
for x in DW2015:
    if x[0] in lifeExpectancy2015:
        drinkingWater2015.append(x)
drinkingWater2015=np.array(drinkingWater2015)
\end{lstlisting}
\end{adjustwidth}

Now, we can see the lists are the same length. If the lengths were still different, we could do the same process with the life expectancy list, i.e. making sure every country in the life expectancy list is also in the drinking water list and removing elements that are not. Next, we want to assign $x$ to the percentages of people with access to basic drinking water services and $y$ to the life expectancies in 2015.
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
x2015=drinkingWater2015[:,2]
x2015=x2015.astype('float64')
y2015=lifeExpectancy2015[:,3]
y2015=y2015.astype('float64')
\end{lstlisting}
\end{adjustwidth}

Now that we have standardized data across the board, we can plot the life expectancy and basic drinking water.

\vspace{-130pt}

\begin{center}
\includegraphics[width=6in]{figure3.pdf}\\
\end{center}

\vspace{-160pt}

\begin{center}
    \textit{Figure 1.}
\end{center}

\vspace{20pt}

Now we create our data matrix $x$, which is the $x$ array transformed into a column vector and appended to a column of ones.
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
x2015=x2015.reshape(-1,1)
X2015=np.concatenate([x2015,np.ones(shape=(x2015.shape[0],1))],axis=1)
\end{lstlisting}
\end{adjustwidth}

Since this is now a column vector, we also want to transform $y$ into a column vector.
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
Y2015=y2015.reshape(-1,1)
\end{lstlisting}
\end{adjustwidth}

Now, we can compute the least squares line and the matrix $M$ that will give us the slope and the $y$-intercept for the line of best fit.
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
D2015=X2015.transpose()@X2015
np.linalg.inv(D2015)
M2015=(np.linalg.inv(D2015)@X2015.transpose())@Y2015
\end{lstlisting}
\end{adjustwidth}

The output is the array $[[0.27859093],[49.15736534]]$, which tells us the slope of the line is 0.279 and the intercept is 49.157. If we want to compare the actual values to predicted values, we use the following code:
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
Yhat2015 = np.dot(np.dot(np.dot(X2015,np.linalg.inv(D2015)),
                                    X2015.transpose()),Y2015)
\end{lstlisting}
\end{adjustwidth}

A visual representation of the least squares line can be seen once executing
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
f2015.line(x=x2015.ravel(),y=Yhat2015[:,0],color='black')
show(f2015)
\end{lstlisting}
\end{adjustwidth}

\vspace{-130pt}

\begin{center}
\includegraphics[width=6in]{figure4.pdf}
\end{center}

\vspace{-160pt}

\begin{center}
    \textit{Figure 2.}
\end{center}

\vspace{20pt}

@@

As seen in \textit{Figure 1}, we can assume a linear regression is appropriate for the model based on the data. There looks to be a linear correlation between the variables rather than a quadratic or other polynomial relationship. To see the R-Squared value of our linear regression, we can use the following code \citep{pythonr2}:
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
corr_matrix = np.corrcoef(Y2015[:,0], Yhat2015[:,0])
corr = corr_matrix[0,1]
R_sq = corr**2
\end{lstlisting}
\end{adjustwidth}
We see the $R^2$ value is 0.66348. So, 66.348\% of the variability in the data is explained by the model. While this is on the lower end, our model is still supported. We would also like to check that the mean of the residuals is close to zero. This can be calculated with the following code.
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
residuals=Yhat2015[:,0]-Y2015[:,0]
np.mean(residuals)
\end{lstlisting}
\end{adjustwidth}
We see the mean is -1.2813065726821479e-14 which is very close to zero. We also need to check for homosteadasticity among the residuals. To look at the residuals, the difference between the measured and predicted values, we can use the following function as detailed in \citet{teitelbaum2021linreg}.
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
def comparison_plot(x,Y, Yhat):
    '''Plots Predicted vs True values for analysis of regression'''
    comparison_plot = 
        figure(title='Difference between Measured and Predicted Values')
    comparison_plot.xaxis.axis_label='x'
    comparison_plot.yaxis.axis_label='Yhat-Y'
    comparison_plot.scatter(x=x,y=Yhat-Y)
    comparison_plot.line(x=[x.min(),x.max()],y=[0,0])
    return comparison_plot
show(comparison_plot(x2015.ravel(),Y2015[:,0], Yhat2015[:,0]))
\end{lstlisting}
\end{adjustwidth}

\vspace{-40pt}

\begin{center}
\includegraphics[width=4in]{figure5.pdf}
\end{center}

\vspace{-50pt}

\begin{center}
    \textit{Figure 3.}
\end{center}

We see there are no patterns among the residuals, so this assumption is validated.

\vspace{15pt}

@@

\begin{adjustwidth}{0.5in}{0.5in}
\begin{lstlisting}
p = sns.displot(residuals,kde=True)
\end{lstlisting}
\end{adjustwidth}

\vspace{-150pt}

\begin{center}
    \includegraphics[width=6in]{figure6.pdf}
\end{center}

\vspace{-160pt}

\begin{center}
\textit{Figure 4.}
\end{center}

\vspace{15pt}

We see the graph is slightly bimodal but almost normally distributed as it is very hard to get perfectly normally distributed curves with real life.

\vspace{15pt}

@@

\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}
df = pd.DataFrame({'x': x2015[:,0],
                   'y': Y2015[:,0]})
model = ols('x ~ y', data=df).fit()
stud_res = model.outlier_test()

for x in stud_res['student_resid']:
    if x > 3 or x < -3:
        print(x)
\end{lstlisting}
\end{adjustwidth}

We see there is one outlier, with a studentized residual of -3.160. This outlier should be removed from the data set.

@@

\vspace{15pt}

This research paper focuses on investigating the effect of a population's access to clean drinking water on their life expectancy. We have found a positive correlation between the variables, which supports existing research that clean water is essential for the body to function. Insights in this field can further promote the basic right to safe water and highlight just how important water is in human health.

>>>>

\documentclass[12pt]{article}

\usepackage{amsmath}
\usepackage[margin=1 in]{geometry}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{natbib}
\usepackage[colorlinks=true, citecolor=blue]{hyperref}
\usepackage{listings}
\usepackage{setspace}
\usepackage{hyperref}

\parindent=0pt
\parskip=\bigskipamount
\onehalfspacing

\title{Using Linear Regression to Examine the Correlation Between Access to Drinking Water and Life Expectancy}
\author{Julia Andronowitz\\B.S., Mathematics and Statistics}
\date{December 9, 2022}

@@

\begin{lstlisting}[language=Python]
    import numpy as np
    import pandas as pd
    import seaborn as sns
    from bokeh.plotting import figure
    from bokeh.io import show, output_notebook
    from bokeh.models import ColumnDataSource, HoverTool
    from bokeh.layouts import gridplot
    output_notebook()
\end{lstlisting}

Next, we import the data:
\begin{lstlisting}[language=Python]
    rawData1=pd.read_csv('archive/basicDrinkingWaterServices.csv')
    rawData2=pd.read_csv('archive/lifeExpectancyAtBirth.csv')
\end{lstlisting}

If we would like to see the data, use the command
\begin{lstlisting}[language=Python]
    rawData1
\end{lstlisting}

An output of the data is in Figures ~\ref{fig1} and ~\ref{fig2}.
\begin{figure}
    \centering
    \includegraphics[width=5in]{Figures/data1.pdf}
    \caption{Access to Basic Drinking Water Services Data}
    \label{fig1}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=5in]{Figures/data2.pdf}
    \caption{Life Expectancy at Birth Data}
    \label{fig2}
\end{figure}
For the drinking water data, we take only the columns with the country, year, and the percentage of the population using at least basic drinking water services.
\begin{lstlisting}[language=Python]
    rawData1=rawData1[["Location", "Period" , "First Tooltip"]]
\end{lstlisting}

We do the same for the life expectancy data, only taking the columns with the country, year, sex, and average life expectancy at birth.
\begin{lstlisting}[language=Python]
    rawData2=rawData2[["Location", "Period", "Dim1", "First Tooltip"]]
\end{lstlisting}

Next, we want to convert our pandas dataframes into numpy arrays so that we can use the appropriate mathematical functions.
\begin{lstlisting}[language=Python]
    data1=np.array(rawData1)
    data2=np.array(rawData2)
\end{lstlisting}

We want to have as many constants as possible in our model. Since the data was collected starting in 2000, we must use the same years in both data sets. We will use the year 2015 as the benchmark. The data can then be filtered using a loop and appending the 2015 data to a new list.
\begin{lstlisting}[language=Python]
    dw2015=[]
    for x in data1:
        if x[1]==2015:
            dw2015.append(x)
\end{lstlisting}

This must be done for both data sets. We now do it for the life expectancy data set. As previously mentioned in the "Data" section, we also want to use only the rows with both sexes.
\begin{lstlisting}[language=Python]
    le2015=[]
    for x in data2:
        if (x[1]==2015) and (x[2]=='Both sexes'):
            le2015.append(x)
\end{lstlisting}

These lists will also be converted into numpy arrays.
\begin{lstlisting}[language=Python]
    DW2015=np.array(dw2015)
    lifeExpectancy2015=np.array(le2015)
\end{lstlisting}

Now, we see the shapes of both these arrays are different. The drinking water array contains 193 rows and 3 columns while the life expectancy array contains 183 rows and 4 columns. This means that there are some countries that are not in both lists. So, we need to delete the data for countries that are only in one or the other. We do this by checking if the countries in the drinking water array are also in the life expectancy data. If both countries are in both data sets, we append them to our new drinking water list, which is what we will use going forward.
\begin{lstlisting}[language=Python]
    drinkingWater2015=[]
    for x in DW2015:
        if x[0] in lifeExpectancy2015:
            drinkingWater2015.append(x)
    drinkingWater2015=np.array(drinkingWater2015)
\end{lstlisting}

Now, we can see the lists are the same length. If the lengths were still different, we could do the same process with the life expectancy list, i.e. making sure every country in the life expectancy list is also in the drinking water list and removing elements that are not. Next, we want to assign $x$ to the percentages of people with access to basic drinking water services and $y$ to the life expectancies in 2015.
\begin{lstlisting}[language=Python]
    x2015=drinkingWater2015[:,2]
    x2015=x2015.astype('float64')
    y2015=lifeExpectancy2015[:,3]
    y2015=y2015.astype('float64')
\end{lstlisting}

Now that we have standardized data across the board, we can plot the life expectancy and basic drinking water. The output is shown in Figure ~\ref{fig3}.
\begin{figure}
    \centering
    \includegraphics[width=4in]{Figures/figure1.png}
    \caption{Life Expectancy vs. Drinking Water}
    \label{fig3}
\end{figure}

Now we create our data matrix $x$, which is the $x$ array transformed into a column vector and appended to a column of ones.
\begin{lstlisting}[language=Python]
    x2015=x2015.reshape(-1,1)
    X2015=np.concatenate([x2015,np.ones(shape=(x2015.shape[0],1))],axis=1)
\end{lstlisting}

Since this is now a column vector, we also want to transform $y$ into a column vector.
\begin{lstlisting}[language=Python]
    Y2015=y2015.reshape(-1,1)
\end{lstlisting}

Now, we can compute the least squares line and the matrix $M$ that will give us the slope and the $y$-intercept for the line of best fit.
\begin{lstlisting}[language=Python]
    D2015=X2015.transpose()@X2015
    np.linalg.inv(D2015)
    M2015=(np.linalg.inv(D2015)@X2015.transpose())@Y2015
\end{lstlisting}

The output is the array $[[0.27859093],[49.15736534]]$, which tells us the slope of the line is 0.279 and the intercept is 49.157. If we want to compare the actual values to predicted values, we use the following code:
\begin{lstlisting}[language=Python]
    Yhat2015 = np.dot(np.dot(np.dot(X2015,np.linalg.inv(D2015)),
                                    X2015.transpose()),Y2015)
\end{lstlisting}

A visual representation of the least squares line can be seen once executing
\begin{lstlisting}[language=Python]
    f2015.line(x=x2015.ravel(),y=Yhat2015[:,0],color='black')
    show(f2015)
\end{lstlisting}
The output is shown in Figure ~\ref{fig4}.
\begin{figure}
    \centering
    \includegraphics[width=4in]{Figures/figure2.png}
    \caption{Plot with Least Squares Line}
    \label{fig4}
\end{figure}

@@

As seen in Figure ~\ref{fig3}, we can assume a linear regression is appropriate for the model based on the data. There looks to be a linear correlation between the variables rather than a quadratic or other polynomial relationship. To see the R-Squared value of our linear regression, we can use the following code \citep{pythonr2}:
\begin{lstlisting}[language=Python]
    corr_matrix = np.corrcoef(Y2015[:,0], Yhat2015[:,0])
    corr = corr_matrix[0,1]
    R_sq = corr**2
\end{lstlisting}
We see the $R^2$ value is 0.66348. So, 66.348\% of the variability in the data is explained by the model. While this is on the lower end, our model is still supported. We would also like to check that the mean of the residuals is close to zero. This can be calculated with the following code.
\begin{lstlisting}[language=Python]
    residuals=Yhat2015[:,0]-Y2015[:,0]
    np.mean(residuals)
\end{lstlisting}
We see the mean is -1.2813065726821479e-14 which is very close to zero. We also need to check for homosteadasticity among the residuals. To look at the residuals, the difference between the measured and predicted values, we can use the following function as detailed in \citet{teitelbaum2021linreg}.
\begin{lstlisting}[language=Python]
    def comparison_plot(x,Y, Yhat):
        '''Plots Predicted vs True values for analysis of regression'''
        comparison_plot = 
            figure(title='Difference between Measured and Predicted Values')
        comparison_plot.xaxis.axis_label='x'
        comparison_plot.yaxis.axis_label='Yhat-Y'
        comparison_plot.scatter(x=x,y=Yhat-Y)
        comparison_plot.line(x=[x.min(),x.max()],y=[0,0])
        return comparison_plot
    show(comparison_plot(x2015.ravel(),Y2015[:,0], Yhat2015[:,0]))
\end{lstlisting}

\begin{figure}
    \centering
    \includegraphics[width=4in]{Figures/figure3.png}
    \caption{Residuals Plot}
    \label{fig5}
\end{figure}

In Figure ~\ref{fig5}, we see there are no patterns among the residuals, so this assumption is validated.

@@

\begin{lstlisting}[language=Python]
    p = sns.displot(residuals,kde=True)
\end{lstlisting}

\begin{figure}
    \centering
    \includegraphics[width=4in]{Figures/figure4.png}
    \caption{Distribution Plot}
    \label{fig6}
\end{figure}

We see the graph in Figure ~\ref{fig6} is slightly bimodal but almost normally distributed as it is very hard to get perfectly normally distributed curves with real life.

@@

\begin{lstlisting}[language=Python]
    df = pd.DataFrame({'x': x2015[:,0],
                       'y': Y2015[:,0]})
    model = ols('x ~ y', data=df).fit()
    stud_res = model.outlier_test()

    for x in stud_res['student_resid']:
        if x > 3 or x < -3:
            print(x)
\end{lstlisting}

Running the code, we see there is one outlier with a studentized residual of -3.160. This outlier should be removed from the data set.

@@

Studies regarding drinking water have been done to draw conclusions about the presence of certain diseases when clean drinking water is limited. Research revolves around specific diseases or access to clean drinking water based on a scale. This research paper focuses on investigating the effect of the proportion of a population's access to clean drinking water on their life expectancy - a more widespread conclusion than those regarding specific water-borne illnesses. We have found a positive correlation between the variables, which supports existing research that clean water is essential for the body to function, yet expands the relationship to overall life expectancy. Insights in this field can further promote the basic right to safe water and highlight just how important water is in human health.
